### 2020-07-20 code/figure1.Rmd
Set up code for figure1b data  section and plotting section

@@

>>>>

####################
## Figure 1B data ##
####################
gdc_path <- gdc_reactome_path_analysis %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample") %>% ## move path_names to column
  .[match(rownames(tumor_samples), .$sample), ]

gdc_path_mca <- left_join(gdc_path, tumor_samples[, c(1, 3)],
  by = c("sample" = "submitted_sample_id")) %>%
  column_to_rownames(var = "sample")

gdc_path_sample_names <- rownames(gdc_path_mca)
gdc_path_tumor_types <- gdc_path_mca$project_code.x

gdc_path_mca$project_code.x <- NULL

## create a categorical variable from the boolean variable
pathway_list <- make_clean_names(colnames(gdc_path_mca))
pathway_count <- length(pathway_list)
gdc_path_mca <- as.data.frame(lapply(gdc_path_mca, function(x) {
  factor(x, levels = c(0, 1), labels = c("no", "yes"))
}))
colnames(gdc_path_mca) <- pathway_list
gdc_path_mca$project_code <- gdc_path_tumor_types
rownames(gdc_path_mca) <- gdc_path_sample_names

## perform mca
mca_res <- MCA(X = gdc_path_mca[, pathway_list], graph = FALSE)

expl_var <- mca_res$eig[, 2]
expl_var_perm <- matrix(NA, ncol = length(mca_res$eig[, 2]), nrow = 100)

set.seed(4242)
for (k in seq(100)) {
  gdc_perm <- as.data.frame(lapply(gdc_path_mca[, pathway_list], sample))
  MCA_perm <- MCA(X = gdc_perm, graph = FALSE)
  expl_var_perm[k, ] <- MCA_perm$eig[, 2]
}

randomize <- tibble(
  eigenvalue = 1:length(expl_var),
  expl_var = expl_var,
  perm = colMeans(expl_var_perm),
  pvalue = rowSums(t(expl_var_perm) >= expl_var) / 100
)

optimal_mca_dimensions <- randomize %>%
  filter(pvalue < 0.05) %>%
  filter(eigenvalue == max(eigenvalue)) %>%
  pull(eigenvalue)

mca_res <- MCA(
  X = gdc_path_mca[, pathway_list],
  ncp = optimal_dimensions, graph = FALSE
)

save(randomize, mca_res, file = "figure1B_data.rda")

@@

optimal_mca_dimensions <- randomize %>%
  filter(pvalue < 0.05) %>%
  filter(eigenvalue == max(eigenvalue)) %>%
  pull(eigenvalue)

permut_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
  geom_line(aes(colour = "#998ec3")) +
  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
  scale_color_identity(guide = guide_legend(title = ""), 
                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
  scale_y_continuous(labels = scales::percent) +
  xlab("MCA Eigenvalues") +
  ylab("explained variance") +
  theme_classic()

pvalue_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = pvalue)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = optimal_dimensions, color = "red") +
  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
  xlab("MCA Eigenvalues") +
  ylab("P-value") +
  theme_classic()

fig1B <- permut_p /
  pvalue_p + plot_annotation(title = "Significance of MCA Eigenvalues")
